A Multiple-Set Overlapping-Domain Decomposed Monte Carlo Synthetic Acceleration Method for Linear Systems

### 2013-09-06 doc/sna2013/paper/slattery_evans_wilson.tex
fixing figure layout in sna paper

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{graphicx}  % allows inclusion of graphics
\usepackage{booktabs}  % nice rules (thick lines) for tables
\usepackage{microtype} % improves typography for PDF
\usepackage{tabls}
\usepackage{afterpage}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage[mathcal]{euscript}
\usepackage{tmath}
\usepackage[section]{placeins}

\usepackage[breaklinks=true,colorlinks=true,linkcolor=black,citecolor=black]{hyperref}

@@

As leadership class machines move beyond the petascale, new algorithms
must be developed that leverage their strengths and adapt to their
shortcomings. Basic research is required now to advance methods in
time for these new machines to become operational. As machines begin
to operate at hundreds of petaflops peak performance and beyond,
trends toward reduced energy consumption will require incredibly high
levels of concurrency to achieve the desired computation rates
\cite{kogge_using_2011}. The end result of these hardware changes is
that the larger number of low-powered processors will be prone to both
soft failures such as bit errors in floating point operations and hard
failures where data cannot be recovered. Because these failures are
predicted to be common, resilient solver technologies are required to
overcome these events at the application level. With linear solvers
based on Monte Carlo techniques, such issues are potentially
alleviated by statistical arguments. In the case of soft failures,
isolated floating point errors in the Monte Carlo simulation are
absorbed within tally statistics. Completely losing memory during a
hard failure is treated as a high variance event where some portion of
the Monte Carlo histories and subsequently the solution are lost with
other histories maintained by replicating the problem and combining
the results through superposition.

In this paper, we will present the Monte Carlo methods of Neumann and
Ulam and then outline Monte Carlo Synthetic Acceleration methods. We
then present a parallel algorithm for MCSA based on the work of
Brunner and Brantley \cite{brunner_efficient_2009} and Wagner
et. al. \cite{wagner_hybrid_2010}. Using this algorithm, we perform a
series of scaling studies to characterize parallel
performance. Finally, we summarize the results and offer future work
that may improve the algorithm.

@@

\begin{figure}[h!]
  \begin{center}
    \scalebox{0.7}{ \input{domain_to_domain.pdftex_t} }
  \end{center}
  \caption{Nearest neighbor history communication sequence.
    \textit{Each subdomain in the system has a set of nearest
      neighbors determined by the parallel adjacency graph of the
      input matrix. The subdomains are indexed by an integer.}}
  \label{fig:nearest_neighbor_comm}
\end{figure}

Given the mutual exclusivity of the transport algorithm, an additional
communcation sequence is required to efficiently stop the algorithm
after all histories have been computed. To stop transport, Brunner and
Brantley use an asynchronous binary communication tree as presented
for the same 9 subdomain example in
Figure~\ref{fig:binary_comm_tree}. In this communication pattern, each
process has one parent process (except for process 0) to which it will
non-blocking send the number of histories that terminated their
transport procedure by weight cutoff in the local
domain. Equivalently, each process has up to two child processes from
which it will receive their completed number of histories with a
non-blocking receive operation. Setting up a tree in this manner lets
the completed history tally be updated incrementally and funneled to
the root process in a mutually exclusive manner. Once the root process
has tallied a completed number of histories equal to the number of
histories in the source, transport is complete. Once this occurs, the
root process non-blocking sends the stop message to its children and
each child process non-blocking receives a stop message from the root
as shown in Figure~\ref{fig:binary_comm_tree}. The stop message is
then propagated up the tree in the same manner between parents and
children.
\begin{figure}[h!]
  \begin{center}
    \scalebox{0.6}{
      \input{binary_comm_tree.pdftex_t} }
  \end{center}
  \caption{Binary communication tree for coordinating the end of a
    parallel Neumann-Ulam solve. \textit{Each child process reports to
      its parents how many histories completed within its domain. When
      the root process sums all complete histories, it forwards the
      stop signal to its children which in turn forward the message up
      the tree. The subdomains are indexed by an integer.}}
  \label{fig:binary_comm_tree}
\end{figure}

@@

To demonstrate the algorithm, consider the example presented in
Figure~\ref{fig:msod_example} adapted from Mervin's work with Wagner
and others in the same area \cite{mervin_variance_2012}.
\begin{figure}[h!]
  \begin{center}
    \scalebox{0.8}{ \input{msod_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Overlapping domain example illustrating how domain
      overlap can reduce communication costs.}
    \textit{All particles start in the blue region of interest. The
      dashed line represents 0.5 domain overlap between domains.}}
  \label{fig:msod_example}
\end{figure}
In this example, 3 particle histories are presented emanating from the
blue region of interest. Starting with particle A, if no domain
overlap is used then the only the blue domain exists on the starting
processor. Particle A is then transported through 3 other domains
before the history ends, therefore requiring three communications to
occur in Brunner's algorithm. If a 0.5 domain overlap is permitted as
shown by the dashed line, then the starting process owns enough of the
domain such that no communications must occur in order to complete the
particle A transport process. Using 0.5 domain overlap also easily
eliminates cases such as that represented by the path of particle
C. In this case, particle C is scattering between two adjacent
domains, incurring a large latency cost for a single
particle. Finally, with particle B we observe that 0.5 domain overlap
will still not eliminate all communications. However, if 1 domain
overlap were used, the entire geometry shown in
Figure~\ref{fig:msod_example} would be contained on the source
processor and therefore transport of all 3 particles without
communication would occur.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Parallel Scaling Studies}
We will next measure performance of the parallel MCSA algorithm in
various scaling studies to assess its quality using a simple
two-dimensional neutron diffusion problem in a homogeneous medium. For
the scaling results presented here, all parallel MCSA results were
computed using the Monte Carlo Linear Solvers\footnote{MCLS is an open
  source C++ library built on the Trilinos\cite{heroux_overview_2005}
  framework and can be downloaded at
  \url{https://github.com/sslattery/MCLS}} (MCLS) library generated
specifically for this work. For every scaling study presented, the
calculations were performed 3 times with the average values reported
for wall time and those average values used for the efficiency
computations.

For all cases, both strong and weak scaling studies were
performed. For strong scaling, the objective is to then decrease the
runtime of a fixed size global problem as a linear function of
$P$. Based on this, we can define the strong scaling \textit{absolute
  efficiency} as \cite{keyes_how_1999}:
\begin{equation}
  \eta_{strong}(N,P) = \frac{1}{P} \frac{T(N,1)}{T(N,P)}\:,
  \label{eq:strong_scaling_absolute}
\end{equation}
where $T(N,P)$ is the wall time for a computation of global size $N$
using $P$ processes. Note here that this particular definition is
measured with respect to a serial computation. On leadership class
machines, it is often the case that a problem size large enough to
effectively execute on $O(10,000)$ cores or larger is significantly
larger than any problem that may be solved by a single core due to
memory restrictions. Therefore, we consider absolute scaling for a
base case where $Q$ processes are used instead of a serial
computation:
\begin{equation}
  \eta_{strong}(N,P|Q) = \frac{Q}{P} \frac{T(N,Q)}{T(N,P)}\:.
  \label{eq:strong_scaling_absolute_ref}
\end{equation}
For weak scaling, the ratio $N/P$ is fixed such that the local problem
size remains the same for all values of $P$. In this case, the
objective is to maintain a constant wall time while both $N$ and $P$
grow as a fixed ratio. In this case, we define the absolute efficiency
as \cite{keyes_how_1999}:
\begin{equation}
  \eta_{weak}(M|N,P|Q) = \frac{T(M,Q)}{T(N,P)}\:,
  \label{eq:weak_scaling_absolute}
\end{equation}
subject to the constraint $M/Q = N/P$. The scaling is perfect if the
runtime is static for all $P$ when $N/P$ is fixed.

The absolute efficiency measures presented here only considers the
total time to solution. Often, that total time to solution may be
changing for an iterative method due to the fact that the number of
iterations required to converge may be changing as a function of
$P$. This will often happen in cases where modifying $N$ changes the
spectral character of the problem, often making it stiffer and
therefore more difficult to solve when $N$ increases. Therefore we
also consider \textit{algorithmic efficiency} which considers this
potential change in iteration count \cite{keyes_how_1999}:
\begin{equation}
  \eta_{alg}(N,P|Q) = \frac{I(N,Q)}{I(N,P)}\:,
  \label{eq:algorithmic_efficiency}
\end{equation}
where $I(N,P)$ is the number of iterations required to converge a
problem of global size $N$ computed using $P$ processes.  All parallel
scaling studies presented in this work were performed on the Titan
Cray XK7 machine at the Oak Ridge Leadership Computing Facility
located at Oak Ridge National Laboratory.

\subsection{Pure Domain Decomposition Scaling}
\label{subsec:pure_domain_decomp}
For the first set of scaling studies, we consider the parallel MCSA
algorithm where MSOD has not been used in the parallel Neumann-Ulam
algorithm. We refer to this case as pure domain decomposition where no
overlap or replication has been leveraged. Such a test will isolate
specifically the performance of Brunner and Brantley's algorithm as
applied to the Neumann-Ulam algorithm. For the strong scaling study,
the global diffusion problem size was fixed at \sn{1.6}{7} DOFs for
all solvers and the 16 core run used as the base
case. Figure~\ref{fig:titan_pure_strong_time} gives the wall time per
iteration for these calculations while
Figure~\ref{fig:titan_pure_strong} gives the results of this study
with the given absolute efficiencies computed using
Eq~(\ref{eq:strong_scaling_absolute}). All MCSA calculations converged
in 22 iterations.

\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=3in]{titan_pure_strong_time.pdf}
  \end{center}
  \caption{Pure domain decomposition strong scaling wall time per
    iteration.}
  \label{fig:titan_pure_strong_time}
\end{figure}

\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=3in]{titan_pure_strong.pdf}
  \end{center}
  \caption{Pure domain decomposition strong scaling absolute
    efficiency. \textit{Super-linear speed-up is from memory thrashing
      in the base case.}}
  \label{fig:titan_pure_strong}
\end{figure}

There are several key features of
Figure~\ref{fig:titan_pure_strong_time} and
Figure~\ref{fig:titan_pure_strong} worth noting. First, up to around
5,000 cores, all methods experience a super-linear improvement in
parallel efficiencies. For all methods, this increase is not a
function of the algorithmic efficiency as those computed using
Eq~(\ref{eq:algorithmic_efficiency}) remained at unity, meaning that
all computations converged in the same number of iterations relative
to the base case. Instead, what is happening here and was observed by
Brunner and Brantley in their work is that the large problem used to
scale to O(10,000) cores was large enough that memory thrashing was
occurring in the 16 core base case. As the number of cores in the
strong scaling study was increased and the local problem size
decreased, more of the local problem fit into the cache of the system,
thus causing improved serial performance and therefore enhanced run
times.

Beyond around 5,000 cores, the global problem size selected for this
strong scaling study causes the local problem size to shrink enough
that parallel communication costs begin to overtake arithmetic costs
on process. This takeover causes this effective strong scaling wall
that is observed in all methods. Using any more cores to solve this
particular global problem is not an efficient use of those parallel
resources. Furthermore, the fact that we see very similar trends in
efficiencies for all methods means that they are bound by the same
parallel performance limitations in a strong scaling environment. This
is important as we can expect the same qualitative scaling performance
from MCSA if arithmetic optimization has been performed to improve the
serial performance of the implementation. In addition, these results
are valuable in that they show MCSA, when properly parallelized, has
the ability to be competitive on leadership-class computing platforms
with conventional subspace methods with good strong scaling observed
in the pure domain decomposition case to O(10,000) cores.

Next, we consider the weak scaling performance of the algorithm. In
this case, we fix the local problem size at \sn{4}{4} DOFs and
increase the number of processors used in the
simulation. Figure~\ref{fig:titan_pure_weak_time} gives the wall time
per iteration and Figure~\ref{fig:titan_weak_absolute} gives the weak
scaling absolute efficiencies for this problem. Overall, weak scaling
performance of the parallel MCSA algorithm is excellent for the pure
domain decomposition case with absolute parallel efficiencies of over
80\% observed at over 32,000 cores.

\begin{figure}[p!]
  \begin{center}
    \includegraphics[width=3in]{titan_pure_weak_time.pdf}
  \end{center}
  \caption{Pure domain decomposition weak scaling wall time
      per iteration.}
  \label{fig:titan_pure_weak_time}
\end{figure}

\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=3in]{titan_weak_absolute.pdf}
  \end{center}
  \caption{\textbf{Pure domain decomposition weak scaling absolute
      efficiency.} \textit{MCLS is an over order of magnitude slower
      arithmetically, causing the higher efficiency. GMRES executes
      more operations than conjugate gradient, also creating a higher
      efficiency.}}
  \label{fig:titan_weak_absolute}
\end{figure}

@@

\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=3in]{titan_strong_overlap.pdf}
  \end{center}
  \caption{\textbf{Strong scaling absolute efficiency with varying
      levels of overlap.} \textit{Values of 0, 1, 2, 3, and 5 levels
      of overlap were used as the problem had a root-mean-squared
      number of states of 2.63 for each history.}}
  \label{fig:titan_strong_overlap}
\end{figure}

\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=3in]{titan_strong_overlap_diff.pdf}
  \end{center}
  \caption{\textbf{Strong scaling absolute efficiency difference with
      respect to the 0 overlap base case.} \textit{Values of 1, 2, 3,
      and 5 levels of overlap are compared here with the absolute
      difference of the absolute efficiencies presented.}}
  \label{fig:titan_strong_overlap_diff}
\end{figure}

@@

\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=3in]{titan_weak_overlap.pdf}
  \end{center}
  \caption{\textbf{Weak scaling absolute efficiency with varying
      levels of overlap.} \textit{Values of 0, 1, 2, 3, and 5 levels
      of overlap were used as the problem had a root-mean-squared
      number of states of 2.63 for each history.}}
  \label{fig:titan_weak_overlap}
\end{figure}

\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=3in]{titan_weak_overlap_diff.pdf}
  \end{center}
  \caption{\textbf{Weak scaling absolute efficiency difference with
      respect to the 0 overlap base case.} \textit{Values of 1, 2, 3,
      and 5 levels of overlap are compared here with the absolute
      difference of the absolute efficiencies presented.}}
  \label{fig:titan_weak_overlap_diff}
\end{figure}

@@

For the multiple sets problem, the single set strong scaling study
from the pure domain decomposition base case was repeated for multiple
sets with both splitting of histories across sets and full replication
of histories across sets using both 2 and 4
sets. Figure~\ref{fig:titan_strong_ms_time} gives the actual wall
times for these computations. There are two important features to note
in the wall time data. First, splitting the histories among sets
instead of replicating them is faster due to the fact that cutting the
number of histories in half reduces the time of an iteration by a
factor that is nearly linear while increasing linearly the number of
histories through replication does not decrease the iterations
required to converge by a linear factor. Second, until the strong
scaling wall is hit starting at around 10,000 cores, the single set
case actually performs the fastest with only marginal improvement in
run times at higher numbers of cores with more sets. At these larger
numbers of cores, however, the improvements are so modest that one is
better off running the same problem using a single set at lower core
counts to conserve computational resources while maintaining a good
time to solution.

\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=3in]{titan_strong_ms_time.pdf}
  \end{center}
  \caption{\textbf{Wall time in seconds to solution for each case for
      the strong scaling study with multiple sets.} \textit{Until the
      strong scaling wall is reached at O(10,000) cores, the single
      set case performs the best. For the multiple set cases, it is
      faster to split histories among sets and maintain iterative
      performance rather than replicate histories to improve iterative
      performance.}}
  \label{fig:titan_strong_ms_time}
\end{figure}

@@

\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=3in]{titan_strong_ms_eff.pdf}
  \end{center}
  \caption{\textbf{Multiple set absolute parallel efficiency.}
    \textit{Computed relative to the 16-core 1-set base case. Using a
      single set is the best means by which to improve strong scaling
      up to the strong scaling wall. At higher core counts past where
      the single set case hits the strong scaling wall adding sets can
      provide some improvements in efficiencies.}}
  \label{fig:titan_strong_ms_eff}
\end{figure}

Also of interest here is the effect of splitting histories among sets
versus replicating them on the iterative performance of the
method. Table~\ref{tab:ms_strong_alg_eff} gives the iterations
required to converge and computed algorithmic efficiencies for each of
the cases presented in the strong scaling data. As expected, splitting
histories maintains a fixed global number of histories and therefore
maintained the number of iterations required to converge relative to
the single set base case. When histories were replicated, adding sets
decreased the number of iterations required to converge in a nonlinear
fashion. Therefore, increasing the number of histories per MCSA
iteration improves algorithmic efficiency. Using these algorithmic
efficiencies, the implementation efficiencies for the multiple set
cases were computed and are given in
Figure~\ref{fig:titan_strong_ms_impeff}. From these results, we see
that the replicated history cases are even less efficient than in the
absolute case with respect to the performance of a single
iteration. This is due to the fact that doubling the number of
histories does not halve the number of MCSA iterations required for
convergence as the acceleration is bound by the Central Limit
Theorem. As a result, the improved iterative performance from the
replicated cases comes at a higher cost in terms of time to solution
than the split cases.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{graphicx}  % allows inclusion of graphics
\usepackage{booktabs}  % nice rules (thick lines) for tables
\usepackage{microtype} % improves typography for PDF
\usepackage{tabls}
\usepackage{afterpage}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage[mathcal]{euscript}
\usepackage{tmath}
\usepackage{float}
\usepackage[breaklinks=true,colorlinks=true,linkcolor=black,citecolor=black]{hyperref}

@@

As leadership class machines move beyond the petascale, new algorithms
must be developed that leverage their strengths and adapt to their
shortcomings. Basic research is required now to advance methods in
time for these new machines to become operational. As machines begin
to operate at hundreds of petaflops peak performance and beyond,
trends toward reduced energy consumption will require incredibly high
levels of concurrency to achieve the desired computation rates
\cite{kogge_using_2011}. The end result of these hardware changes is
that the larger number of low-powered processors will be prone to both
soft failures such as bit errors in floating point operations and hard
failures where data cannot be recovered. Because these failures are
predicted to be common, resilient solver technologies are required to
overcome these events at the application level. With linear solvers
based on Monte Carlo techniques, such issues are potentially
alleviated by statistical arguments. In the case of soft failures,
isolated floating point errors in the Monte Carlo simulation are
absorbed within tally statistics. Completely losing memory during a
hard failure is treated as a high variance event where some portion of
the Monte Carlo histories and subsequently of the solution are lost
with other histories maintained by replicating the problem and
combining the results through superposition.

In this paper, we will present the Monte Carlo method of Neumann and
Ulam for solving linear systems and then outline Monte Carlo Synthetic
Acceleration. We then present a parallel algorithm for MCSA based on
the work of Brunner and Brantley \cite{brunner_efficient_2009} and
Wagner et. al. \cite{wagner_hybrid_2010}. Using this algorithm, we
perform a series of scaling studies to characterize parallel
performance. Finally, we summarize the results and offer future work
that may improve the algorithm.

@@

\begin{figure}[h!]
  \begin{center}
    \scalebox{0.7}{ \input{domain_to_domain.pdftex_t} }
  \end{center}
  \caption{Nearest neighbor history communication sequence.
    \textit{Each subdomain in the system has a set of nearest
      neighbors determined by the parallel adjacency graph of the
      input matrix. The subdomains are indexed by an integer.}}
  \label{fig:nearest_neighbor_comm}
\end{figure}
Given the mutual exclusivity of the transport algorithm, an additional
communcation sequence is required to efficiently stop the algorithm
after all histories have been computed. To stop transport, Brunner and
Brantley use an asynchronous binary communication tree as presented
for the same 9 subdomain example in
Figure~\ref{fig:binary_comm_tree}. In this communication pattern, each
process has one parent process (except for process 0) to which it will
non-blocking send the number of histories that terminated their
transport procedure by weight cutoff in the local
domain. Equivalently, each process has up to two child processes from
which it will receive their completed number of histories with a
non-blocking receive operation. Setting up a tree in this manner lets
the completed history tally be updated incrementally and funneled to
the root process in a mutually exclusive manner. Once the root process
has tallied a completed number of histories equal to the number of
histories in the source, transport is complete. Once this occurs, the
root process non-blocking sends the stop message to its children and
each child process non-blocking receives a stop message from the root
as shown in Figure~\ref{fig:binary_comm_tree}. The stop message is
then propagated up the tree in the same manner between parents and
children.
\begin{figure}[h!]
  \begin{center}
    \scalebox{0.6}{
      \input{binary_comm_tree.pdftex_t} }
  \end{center}
  \caption{Binary communication tree for coordinating the end of a
    parallel Neumann-Ulam solve. \textit{Each child process reports to
      its parents how many histories completed within its domain. When
      the root process sums all complete histories, it forwards the
      stop signal to its children which in turn forward the message up
      the tree. The subdomains are indexed by an integer.}}
  \label{fig:binary_comm_tree}
\end{figure}

@@

To demonstrate the algorithm, consider the example presented in
Figure~\ref{fig:msod_example} adapted from Mervin's work with Wagner
and others in the same area \cite{mervin_variance_2012}.
\begin{figure}[H]
  \begin{center}
    \scalebox{0.8}{ \input{msod_example.pdftex_t} }
  \end{center}
  \caption{Overlapping domain example illustrating how domain overlap
    can reduce communication costs. \textit{All particles start in the
      blue region of interest. The dashed line represents 0.5 domain
      overlap between domains.}}
  \label{fig:msod_example}
\end{figure}
In this example, 3 particle histories are presented emanating from the
blue region of interest. Starting with particle A, if no domain
overlap is used then the only the blue domain exists on the starting
processor. Particle A is then transported through 3 other domains
before the history ends, therefore requiring three communications to
occur in Brunner's algorithm. If a 0.5 domain overlap is permitted as
shown by the dashed line, then the starting process owns enough of the
domain such that no communications must occur in order to complete the
particle A transport process. Using 0.5 domain overlap also easily
eliminates cases such as that represented by the path of particle
C. In this case, particle C is scattering between two adjacent
domains, incurring a large latency cost for a single
particle. Finally, with particle B we observe that 0.5 domain overlap
will still not eliminate all communications. However, if 1 domain
overlap were used, the entire geometry shown in
Figure~\ref{fig:msod_example} would be contained on the source
processor and therefore transport of all 3 particles without
communication would occur.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Parallel Scaling Studies}
We will next measure performance of the parallel MCSA algorithm in
various scaling studies to assess its quality using a simple
two-dimensional neutron diffusion problem in a homogeneous medium. For
the scaling results presented here, all parallel MCSA results were
computed using the Monte Carlo Linear Solvers\footnote{MCLS is an open
  source C++ library built on the Trilinos\cite{heroux_overview_2005}
  framework and can be downloaded at
  \url{https://github.com/sslattery/MCLS}} (MCLS) library which
contains implementations of the parallel Neumann-Ulam and MCSA
algorithms. For every scaling study presented, the calculations were
performed 3 times with the average values reported for wall time and
those average values used for the efficiency computations. All
parallel scaling studies presented in this work were performed on the
Titan Cray XK7 machine at the Oak Ridge Leadership Computing Facility
located at Oak Ridge National Laboratory.

For all cases, both strong and weak scaling studies were
performed. For strong scaling, the objective is to decrease the
runtime of a fixed size global problem as a linear function of $P$,
where $P$ is the number of processors being used. Based on this, we
can define the strong scaling \textit{absolute efficiency} as
\cite{keyes_how_1999}:
\begin{equation}
  \eta_{strong}(N,P) = \frac{1}{P} \frac{T(N,1)}{T(N,P)}\:,
  \label{eq:strong_scaling_absolute}
\end{equation}
where $T(N,P)$ is the wall time for a computation of global size $N$
using $P$ processes. Note here that this particular definition is
measured with respect to a serial computation. On leadership-class
machines, it is often the case that we use a reference case that
occupies an entire node in the system. Therefore, we consider absolute
scaling for a reference case where $Q$ processes are used instead of a
serial computation:
\begin{equation}
  \eta_{strong}(N,P|Q) = \frac{Q}{P} \frac{T(N,Q)}{T(N,P)}\:.
  \label{eq:strong_scaling_absolute_ref}
\end{equation}
For weak scaling, the ratio $N/P$ is fixed such that the local problem
size remains the same for all values of $P$. In this case, the
objective is to maintain a constant wall time while both $N$ and $P$
grow as a fixed ratio. In this case, we define the absolute efficiency
as \cite{keyes_how_1999}:
\begin{equation}
  \eta_{weak}(M|N,P|Q) = \frac{T(M,Q)}{T(N,P)}\:,
  \label{eq:weak_scaling_absolute}
\end{equation}
subject to the constraint $M/Q = N/P$. The scaling is perfect if the
runtime is static for all $P$ when $N/P$ is fixed.

The absolute efficiency measures presented here only consider the
total time to solution. Often, that total time to solution may be
changing for an iterative method due to the fact that the number of
iterations required to converge may be changing as a function of
$N$. This will often happen in cases where modifying $N$ changes the
spectral character of the problem, often making it stiffer and
therefore more difficult to solve when $N$ increases. Therefore, we
also consider \textit{algorithmic efficiency} which quantifies this
change in iteration count \cite{keyes_how_1999}:
\begin{equation}
  \eta_{alg}(N,P|Q) = \frac{I(N,Q)}{I(N,P)}\:,
  \label{eq:algorithmic_efficiency}
\end{equation}
where $I(N,P)$ is the number of iterations required to converge a
problem of global size $N$ computed using $P$ processes.

\subsection{Pure Domain Decomposition Scaling}
\label{subsec:pure_domain_decomp}
For the first set of scaling studies, we consider the parallel MCSA
algorithm where MSOD has not been used in the parallel Neumann-Ulam
algorithm. We refer to this case as pure domain decomposition where no
overlap or replication has been leveraged. Such a test will
specifically isolate the performance of Brunner and Brantley's
algorithm as applied to the random walk sequence. For the strong
scaling study, the global diffusion problem size was fixed at a
relatively small \sn{1.6}{7} DOFs and the 16 core run used as the
reference case. Figure~\ref{fig:titan_pure_strong_time} gives the wall
time per iteration for these calculations while
Figure~\ref{fig:titan_pure_strong} gives the absolute efficiencies
computed using Eq~(\ref{eq:strong_scaling_absolute}). All MCSA
calculations converged the diffusion problem in 22 iterations.

\begin{figure}[H]
  \begin{center}
    \includegraphics[width=3in]{titan_pure_strong_time.pdf}
  \end{center}
  \caption{Pure domain decomposition strong scaling wall time per
    iteration.}
  \label{fig:titan_pure_strong_time}
\end{figure}

\begin{figure}[H]
  \begin{center}
    \includegraphics[width=3in]{titan_pure_strong.pdf}
  \end{center}
  \caption{Pure domain decomposition strong scaling absolute
    efficiency. \textit{Super-linear speed-up is from memory thrashing
      in the reference case.}}
  \label{fig:titan_pure_strong}
\end{figure}

There are several features of Figure~\ref{fig:titan_pure_strong_time}
and Figure~\ref{fig:titan_pure_strong} worth noting. First, up to
around 5,000 cores we observe a super-linear improvement in parallel
efficiencies. This increase was also observed by Brunner and Brantley
in their work and is a result of using a problem large enough to scale
to O(10,000) cores from 16, generating a large enough memory footprint
that memory thrashing was occurring in the 16 core reference case. As
the number of cores in the strong scaling study was increased and the
local problem size decreased, more of the local problem fit into the
cache of the system, thus causing improved serial performance and
therefore enhanced run times and efficiencies. Beyond around 5,000
cores, the global problem size selected for this strong scaling study
causes the local problem size to shrink enough that parallel
communication costs begin to overtake serial work. This takeover
causes the strong scaling wall that is observed in
Figure~\ref{fig:titan_pure_strong}. Using any more cores to solve this
particular global problem is not an efficient use of those parallel
resources.

Next, we consider the weak scaling performance of the algorithm. In
this case, we fix the local problem size at \sn{4}{4} DOFs and
increase the number of processors used in the
simulation. Figure~\ref{fig:titan_pure_weak_time} gives the wall time
per iteration and Figure~\ref{fig:titan_weak_absolute} gives the weak
scaling absolute efficiencies. Overall, weak scaling performance of
the parallel MCSA algorithm is excellent for the pure domain
decomposition case with absolute parallel efficiencies of over 80\%
observed at over 32,000 cores.

\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=3in]{titan_pure_weak_time.pdf}
  \end{center}
  \caption{Pure domain decomposition weak scaling wall time
      per iteration.}
  \label{fig:titan_pure_weak_time}
\end{figure}

\begin{figure}[H]
  \begin{center}
    \includegraphics[width=3in]{titan_weak_absolute.pdf}
  \end{center}
  \caption{Pure domain decomposition weak scaling absolute
    efficiency. \textit{MCLS is an over order of magnitude slower
      arithmetically, causing the higher efficiency. GMRES executes
      more operations than conjugate gradient, also creating a higher
      efficiency.}}
  \label{fig:titan_weak_absolute}
\end{figure}

@@

\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=3in]{titan_strong_overlap.pdf}
  \end{center}
  \caption{Strong scaling absolute efficiency with varying levels of
    overlap. \textit{Values of 0, 1, 2, 3, and 5 levels of overlap
      were used as the problem had a root-mean-squared number of
      states of 2.63 for each history.}}
  \label{fig:titan_strong_overlap}
\end{figure}

\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=3in]{titan_strong_overlap_diff.pdf}
  \end{center}
  \caption{Strong scaling absolute efficiency difference with
      respect to the 0 overlap reference case. \textit{Values of 1, 2, 3,
      and 5 levels of overlap are compared here with the absolute
      difference of the absolute efficiencies presented.}}
  \label{fig:titan_strong_overlap_diff}
\end{figure}

@@

\begin{figure}[H]
  \begin{center}
    \includegraphics[width=3in]{titan_weak_overlap.pdf}
  \end{center}
  \caption{Weak scaling absolute efficiency with varying levels of
    overlap. \textit{Values of 0, 1, 2, 3, and 5 levels of overlap
      were used as the problem had a root-mean-squared number of
      states of 2.63 for each history.}}
  \label{fig:titan_weak_overlap}
\end{figure}

\begin{figure}[H]
  \begin{center}
    \includegraphics[width=3in]{titan_weak_overlap_diff.pdf}
  \end{center}
  \caption{Weak scaling absolute efficiency difference with respect to
    the 0 overlap reference case. \textit{Values of 1, 2, 3, and 5
      levels of overlap are compared here with the absolute difference
      of the absolute efficiencies presented.}}
  \label{fig:titan_weak_overlap_diff}
\end{figure}

@@

For the multiple sets problem, the single set strong scaling study
from the pure domain decomposition reference case was repeated for
multiple sets with both splitting of histories across sets and full
replication of histories across sets using both 2 and 4
sets. Figure~\ref{fig:titan_strong_ms_time} gives the actual wall
times for these computations. There are two important features to note
in the wall time data. First, splitting the histories among sets
instead of replicating them is faster due to the fact that cutting the
number of histories in half reduces the time of an iteration by a
factor that is nearly linear while increasing linearly the number of
histories through replication does not decrease the iterations
required to converge by a linear factor. Second, until the strong
scaling wall is hit starting at around 10,000 cores, the single set
case actually performs the fastest with only marginal improvement in
run times at higher numbers of cores with more sets. At these larger
numbers of cores, however, the improvements are so modest that one is
better off running the same problem using a single set at lower core
counts to conserve computational resources while maintaining a good
time to solution.

\begin{figure}[H]
  \begin{center}
    \includegraphics[width=3in]{titan_strong_ms_time.pdf}
  \end{center}
  \caption{Wall time in seconds to solution for each case for the
    strong scaling study with multiple sets. \textit{Until the strong
      scaling wall is reached at O(10,000) cores, the single set case
      performs the best. For the multiple set cases, it is faster to
      split histories among sets and maintain iterative performance
      rather than replicate histories to improve iterative
      performance.}}
  \label{fig:titan_strong_ms_time}
\end{figure}

@@

\begin{figure}[H]
  \begin{center}
    \includegraphics[width=3in]{titan_strong_ms_eff.pdf}
  \end{center}
  \caption{Multiple set absolute parallel efficiency.
    \textit{Computed relative to the 16-core 1-set reference
      case. Using a single set is the best means by which to improve
      strong scaling up to the strong scaling wall. At higher core
      counts past where the single set case hits the strong scaling
      wall adding sets can provide some improvements in
      efficiencies.}}
  \label{fig:titan_strong_ms_eff}
\end{figure}

Also of interest here is the effect of splitting histories among sets
versus replicating them on the iterative performance of the
method. Table~\ref{tab:ms_strong_alg_eff} gives the iterations
required to converge and computed algorithmic efficiencies for each of
the cases presented in the strong scaling data. As expected, splitting
histories maintains a fixed global number of histories and therefore
maintained the number of iterations required to converge relative to
the single set reference case. When histories were replicated, adding
sets decreased the number of iterations required to converge in a
nonlinear fashion. Therefore, increasing the number of histories per
MCSA iteration improves algorithmic efficiency. Using these
algorithmic efficiencies, the implementation efficiencies for the
multiple set cases were computed and are given in
Figure~\ref{fig:titan_strong_ms_impeff}. From these results, we see
that the replicated history cases are even less efficient than in the
absolute case with respect to the performance of a single
iteration. This is due to the fact that doubling the number of
histories does not halve the number of MCSA iterations required for
convergence as the acceleration is bound by the Central Limit
Theorem. As a result, the improved iterative performance from the
replicated cases comes at a higher cost in terms of time to solution
than the split cases.

\begin{table}[H]
  \begin{center}
    \begin{tabular}{lcc}\hline\hline
      \multicolumn{1}{l}{Case}& 
      \multicolumn{1}{c}{Iterations}&
      \multicolumn{1}{c}{$\eta_{alg}$} \\\hline
      1 & 22 & 1.0 \\
      2-split & 22 & 1.0 \\
      2-replicated & 16 & 1.375 \\
      4-split & 22 & 1.0 \\
      4-replicated & 13 & 1.692 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{Strong scaling algorithmic efficiency using multiple
    sets. \textit{Replicating histories improves algorithmic
      efficiency by converging in fewer iterations.}}
  \label{tab:ms_strong_alg_eff}
\end{table}
